---
title: "Green Index Calculations by Subregion: Karlova District, Tartu, Estonia"
author: "Jessica Middleton"
date: "March 2026"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

---

# Introduction

This script calculates Green Index values for the historic residential core of Karlova, Tartu, Estonia, loosely following the methodology described in:

> _Urban Neighborhood Green Index – A measure of green spaces in urban areas_
> doi: [10.1016/j.landurbplan.2012.01.003](https://doi.org/10.1016/j.landurbplan.2012.01.003)

**Green Index Definition:** The percentage of greenery within each 100m × 100m grid cell, derived from binary classification of NDVI measurements. Cells are classified into four quality levels on a scale of 0.25 - 1:

| Green Index Value | % Greenness in Cell | Label |
|:-----------------:|:-------------------:|:-----:|
| 0 | < 25% | No Greenness |
| 0.25 | 25% – 49% | Low Greenness |
| 0.50 | 50% – 74% | Medium Greenness |
| 1.00 | ≥ 75% | High Greenness |

> _Note: The four equal interval classification was adopted for simplicity, as this index provides relative values for comparative evaluation between neighbourhoods rather than absolute values._

---

# Workflow Summary

**Software:** RStudio (v2025.09.0-387), R (v4.5.1), QGIS (v3.28)  
**Validation:** Results verified post-analysis in QGIS (v3.28)

**Data sources:**

- NDVI data (TIFF, 10m × 10m resolution): provided by course instructors via Moodle
- Study area polygon (GeoPackage): generated by project groupmate

**Workflow steps:**

1. Load and reproject NDVI raster and study area polygon to EPSG:3857
2. Clip NDVI to study area extent
3. Create 100m × 100m aggregation grid
4. Binary-classify NDVI (green = NDVI > 0.25) and extract mean greenness per cell
5. Assign greenness labels from percentage thresholds
6. Export final masked grid as GeoPackage
7. Statistical testing (Kruskal-Wallis + Dunn's post-hoc) by subregion

**NOTE:** Subregion boundaries were manually delineated in QGIS based on land use polygons. These subregions were joined to each Green Index cell based on location, and the results were saved as a geopackage and imported back into RStudio to continue with statistical testing. 

---

# Libraries

```{r libraries}

library(terra)          # Reading, processing, and analyzing raster and vector spatial data
library(sf)             # Handling vector spatial data
library(exactextractr)  # Extracting raster values within polygons (e.g. NDVI per grid cell)
library(dplyr)          # Data wrangling and manipulation (filtering, grouping, summarizing)
library(ggplot2)        # Creating plots and visualizations (e.g. boxplots)
library(dunn.test)      # Post-hoc Dunn's test following Kruskal-Wallis significance testing
library(multcompView)   # Generating compact letter displays from pairwise comparison results
library(writexl)        # Exporting data frames to Excel (.xlsx) files
library(knitr)          # Table formatting in R Markdown output
library(kableExtra)     # Table formatting in R Markdown output

```

---

# Data Preparation

## Load and Format NDVI Raster

```{r loadNDVI}

# Fix occasional CRS resolution errors with terra library
Sys.setenv(PROJ_LIB = system.file("proj", package = "terra"))
Sys.setenv(PROJ_DATA = system.file("proj", package = "terra"))

# Load NDVI raster and assign CRS
ndvi <- rast("Tartu_NDVI_2024_ver2.tif")
crs(ndvi) <- "epsg:3857"

# Confirm CRS assigned correctly
crs(ndvi, describe = TRUE)

# Visualize to confirm alignment with expected data
plot(ndvi, main = "NDVI: Tartu (2024)")

```

## Load and Format Study Area Polygon

```{r loadStudyArea}

# Load study area polygon and reproject to match NDVI CRS
studyarea <- vect("research_area.gpkg")
studyarea <- project(studyarea, crs(ndvi))

# Confirm CRS alignment
crs(studyarea, describe = TRUE)
print(same.crs(ndvi, studyarea))

# Visualize study area boundary
plot(studyarea, main = "Study Area: Karlova District")

```

## Confirm Initial Alignment

```{r finalInitAlignment}

# Overlay NDVI and study area to confirm spatial alignment before clipping
plot(ndvi, main = "NDVI with Study Area Overlay")
plot(studyarea, add = TRUE, col = "red")

```

## Clip NDVI to Study Area

```{r maskRast}

# Mask NDVI to study area extent
ndvi_masked <- mask(ndvi, studyarea)
ndvi_masked <- project(ndvi_masked, crs(ndvi))

# Confirm CRS consistency across all layers
crs(ndvi_masked, describe = TRUE)
print(same.crs(ndvi, studyarea) & same.crs(ndvi, ndvi_masked))

# Plot masked NDVI with slight margin
bbox         <- st_bbox(studyarea)
x_range      <- bbox["xmax"] - bbox["xmin"]
y_range      <- bbox["ymax"] - bbox["ymin"]
margin_factor <- 0.1

expanded_bbox <- bbox
expanded_bbox["xmin"] <- bbox["xmin"] - margin_factor * x_range
expanded_bbox["xmax"] <- bbox["xmax"] + margin_factor * x_range
expanded_bbox["ymin"] <- bbox["ymin"] - margin_factor * y_range
expanded_bbox["ymax"] <- bbox["ymax"] + margin_factor * y_range

plot(ndvi_masked,
     main = "Masked NDVI: Karlova District",
     xlim = c(expanded_bbox["xmin"], expanded_bbox["xmax"]),
     ylim = c(expanded_bbox["ymin"], expanded_bbox["ymax"]))

```

---

# Calculate Green Index

**Green Index** = Percentage of greenness within each 100m × 100m grid cell

| % Green | Label |
|:-------:|:-----:|
| < 25% | No Greenness |
| 25% – 49% | Low Greenness |
| 50% – 74% | Medium Greenness |
| ≥ 75% | High Greenness |

## Create 100m × 100m Aggregation Grid

```{r createGrid, results='hold'}

# Report original raster dimensions and resolution
cat("Original raster size:", ncol(ndvi_masked), "columns x", nrow(ndvi_masked), "rows\n")
cat("Original resolution: ", xres(ndvi_masked), "x", yres(ndvi_masked), "map units per pixel\n\n")

# Aggregate to 100m x 100m (10x the original 10m resolution)
new_cell_size_x <- xres(ndvi_masked) * 10
new_cell_size_y <- yres(ndvi_masked) * 10

grid    <- st_make_grid(st_as_sfc(st_bbox(ndvi_masked)),
                        cellsize = c(new_cell_size_x, new_cell_size_y),
                        square = TRUE)
grid_sf <- st_sf(id = seq_along(grid), geometry = grid)

cat("New grid size:", nrow(grid_sf), "cells\n")
cat("New cell size:", new_cell_size_x, "x", new_cell_size_y, "map units\n\n")

# Confirm CRS consistency
print(crs(grid_sf) == crs(ndvi_masked))

```

## Extract NDVI Values to Grid

> **Note:** Pixels outside the study area polygon were assigned `NaN` during masking. Cells with no valid NDVI data are excluded from analysis. Accuracy of the masked dataset was confirmed in QGIS during post-processing  

```{r NDVI_to_Grid}

# Binary classification: pixels with NDVI > 0.25 are classified as green
green_pixels <- ndvi_masked > 0.25

# Extract mean greenness per grid cell and convert to percentage
green_percentage       <- exact_extract(green_pixels, grid_sf, 'mean') * 100
grid_sf$percent_green  <- round(green_percentage, 2)

# Remove cells with no valid data
valid_grid <- grid_sf[!is.na(grid_sf$percent_green), ]

print(head(valid_grid))

```

## Assign Greenness Labels

```{r setCategories}

# Classify grid cells into greenness categories based on percentage thresholds
grid_sf$greenness <- cut(
  grid_sf$percent_green,
  breaks = c(0, 25, 50, 75, 100),
  labels = c("No Greenness",
             "Low Greenness",
             "Medium Greenness",
             "High Greenness"),
  include.lowest = TRUE,
  right = FALSE   # intervals are [0,25), [25,50), [50,75), [75,100]
)

```

## Plot Final Green Index and Verify Accuracy

```{r plotFinalVerify}

# Define color scheme for greenness labels
colors <- c(
  "No Greenness" = "brown",
  "Low Greenness" = "darkorange",
  "Medium Greenness" = "gold",
  "High Greenness" = "green"
)

# Compute expanded bounding box for plot context
bbox          <- st_bbox(studyarea)
x_range       <- bbox["xmax"] - bbox["xmin"]
y_range       <- bbox["ymax"] - bbox["ymin"]
margin_factor <- 0.1

expanded_bbox          <- bbox
expanded_bbox["xmin"]  <- bbox["xmin"] - margin_factor * x_range
expanded_bbox["xmax"]  <- bbox["xmax"] + margin_factor * x_range
expanded_bbox["ymin"]  <- bbox["ymin"] - margin_factor * y_range
expanded_bbox["ymax"]  <- bbox["ymax"] + margin_factor * y_range

# Remove NA cells, convert to SpatVector, and mask to study area
non_na_grid_sf   <- grid_sf[!is.na(grid_sf$greenness), ]
non_na_grid_vect <- vect(non_na_grid_sf)
masked_grid_sf   <- mask(non_na_grid_vect, studyarea)

# Plot with greenness color scheme
plot(masked_grid_sf["greenness"],
     col      = colors[as.character(masked_grid_sf$greenness)],
     main     = "Green Index: Karlova District, Tartu",
     cex.main = 1.2,
     xlim = c(expanded_bbox["xmin"], expanded_bbox["xmax"]),
     ylim = c(expanded_bbox["ymin"], expanded_bbox["ymax"]))

```

---
   
# Export Final SpatVector as GeoPackage

```{r geopackageExport}

# Uncomment the line below to export the final masked green index layer
# writeVector(masked_grid_sf, "green_index_masked.gpkg", overwrite = TRUE)

```


The aggregated results from *green_index_masked.gpkg* were imported into QGIS for further spatial processing. Geographic subregions were then manually delineated with reference to land cover vector data. Each grid cell was assigned to a subregion based on polygon containment. A residual category, Subregion 8, was created for grid cells that did not fall within any of the predefined land cover polygons. The updated GeoPackage was then reimported into R for statistical analysis.


---

# Statistical Testing

The following section tests whether greenness levels differ significantly across subregions of the study area.

## Assign Ordinal Green Rank

```{r greenRank}

# Assign ordinal rank (1 = highest greenness, 4 = no greenness) for statistical testing
masked_grid_sf$green_rank <- NA
masked_grid_sf$green_rank[masked_grid_sf$greenness == "High Greenness"]   <- 1
masked_grid_sf$green_rank[masked_grid_sf$greenness == "Medium Greenness"] <- 2
masked_grid_sf$green_rank[masked_grid_sf$greenness == "Low Greenness"]    <- 3
masked_grid_sf$green_rank[masked_grid_sf$greenness == "No Greenness"]     <- 4

head(masked_grid_sf)

```

## Load Green Index Output for Statistical Analysis

```{r loadForStats}

# Load the exported GeoPackage output for statistical analysis
# This file is generated by the export step above and includes subregion attribution
file <- st_read("green_index_output.gpkg")

plot(file, main = "Green Index Output: Subregions")

# Check for missing values before proceeding
cat("Missing log_percgreen values:", sum(is.na(file$log_percgreen)), "\n")
cat("Missing subregion values:",     sum(is.na(file$subregion)), "\n")

```

## Log-Transform Green Percentage

```{r normalizeData}

# Log-transform percent_green to reduce right skew before normality testing
# 0.01 is added to handle zero values
file$log_percgreen <- log(file$percent_green + 0.01)

```

## Test for Normality

```{r testAssumptions}

# Shapiro-Wilk normality test on log-transformed values
# If p < 0.05, data is not normally distributed and non-parametric tests are required
shapiro.test(file$log_percgreen)

```

## Kruskal-Wallis Test

```{r nonparametric}

# Non-parametric equivalent of one-way ANOVA
# Tests whether greenness distributions differ significantly across subregions
krus_test <- kruskal.test(log_percgreen ~ subregion, data = file)
print(krus_test)

```

## Post-Hoc Dunn's Test

```{r posthoc}

# Dunn's test with Bonferroni correction identifies which subregions differ significantly
dunn_res <- dunn.test(file$log_percgreen, file$subregion, method = "bonferroni")

```

## Compact Letter Display and Boxplot

```{r cldBoxplot}

# Order subregion factor levels alphabetically 
file$subregion <- factor(file$subregion, levels = sort(unique(file$subregion)))

# Extract adjusted p-values and format comparison names for multcompLetters
pvals        <- dunn_res$P.adjusted
comparisons  <- dunn_res$comparisons
names(pvals) <- gsub(" vs ", "-", comparisons)

# Generate compact letter display 
cld <- multcompLetters(pvals, threshold = 0.05)

# Build summary dataframe by subregion
summary_df <- file %>%
  group_by(subregion) %>%
  summarize(median = median(log_percgreen, na.rm = TRUE)) %>%
  mutate(letter = cld$Letters[as.character(subregion)])

# Plot boxplot with Dunn's test letters annotated
ggplot(file, aes(x = factor(subregion), y = log_percgreen)) +
  geom_boxplot(alpha = 0.7) +
  geom_text(data = summary_df,
            aes(x = factor(subregion), y = median + 0.1, label = letter),
            colour = "red", size = 5) +
  labs(x     = "Subregion",
       y     = "Log-transformed % Green Cover",
       title = "Green Cover by Subregion: Dunn's Test Post-Hoc Letters") +
  expand_limits(y = max(summary_df$median + 1)) +
  theme_minimal()

```

## Greenness Proportion Table by Subregion

```{r greenConnectivityTable}

# Define the all_levels vector (in case not already in environment)
all_levels <- c("No Greenness", "Low Greenness", "Medium Greenness", "High Greenness")

# Create proportion_table 
proportion_table <- file %>%
  st_drop_geometry() %>%
  filter(!is.na(green_connectivity)) %>%
  mutate(green_connectivity = factor(green_connectivity, levels = all_levels)) %>%
  count(subregion, green_connectivity, .drop = FALSE) %>%
  group_by(subregion) %>%
  mutate(percentage = round((n / sum(n)) * 100, 2)) %>%
  ungroup()

# Define colors for the 8 subregions
subregion_colors <- c(
  "Subregion 1" = "#D1E9F6", "Subregion 2" = "#F6EACB", 
  "Subregion 3" = "#D1D8C5", "Subregion 4" = "#EADBC8", 
  "Subregion 5" = "#F1D4E5", "Subregion 6" = "#E1AFD1", 
  "Subregion 7" = "#FFE6E6", "Subregion 8" = "#E5E1DA"
)

# Generate and style table
proportion_table %>%
  mutate(
    subregion = cell_spec(subregion, "html", 
                          background = subregion_colors[subregion],
                          bold = TRUE,
                          color = "#333333")
  ) %>%
  kable(
    format = "html", 
    escape = FALSE, 
    caption = "Greenness Classification by Subregion",
    col.names = c("Subregion", "Greenness Classification", "Count", "Percentage (%)"),
    align = "llrr"
  ) %>%
  kable_styling(
    bootstrap_options = c("hover", "condensed"),
    full_width = TRUE
  ) %>%
  column_spec(1, width = "20%", extra_css = "padding-left: 20px; padding-right: 20px;") %>%
  column_spec(2, width = "30%", extra_css = "padding-left: 20px; padding-right: 20px;") %>%
  column_spec(3:4, width = "25%", extra_css = "padding-right: 30px;") %>%
  collapse_rows(columns = 1, valign = "middle")

```
